---
title: data cleaning
---

```{r libraries}
library(tidyverse)
library(explore)
```


## Total Space

```{r data}

df1 <- read.csv("./data4/absolute_counting_colabs_organic.csv")
df1 |> explore()
```

```{r explore}

df1.2 <- df1 |> 
    pivot_longer(
        cols = starts_with("X"),
        names_to = "year",
        values_to = "value_raw",
        values_drop_na = TRUE
  )

df1.2 |> explore()
```

```{r transform}

df1.3 <- df1.2 %>%
  group_by(year) %>%
  mutate(percentage = (value_raw / value_raw[Country == "Total"]) * 100) %>%
  ungroup()

df1.3 |> explore()
```

::: {.callout-note}
    group_by(year): Groups the data by the year column.

    mutate(percentage = ...): Creates the new column percentage.

    value_raw[Country == "Total"]: Retrieves the total value for the corresponding year.

    Each country's value is divided by its year's total and multiplied by 100 to get the percentage.

    The Total rows will naturally calculate to 100% since they're divided by themselves.
:::

```{r clean}

library(dplyr)

# Step 1: Remove rows with NA or null in the percentage column
df1.4 <- df1.3 %>%
  filter(!is.na(percentage))

# Step 2: Clean the year column
df1.5 <- df1.4 %>%
  mutate(year = as.numeric(gsub("X", "", year))) |>
  filter(!Country == "Total") |> 
  filter(year >= 1996)
# View the cleaned dataframe
df1.5 |> explore()
df1.5 |> glimpse()

write.csv(df1.5, "./data5/Organometallic.csv")
```


```{r country_codes}
country_codes <- read.csv("./data4/country_codes.csv")
```


```{r}
tictoc::tic()
library(readxl)
library(writexl)
library(dplyr)
library(stringr)

# Read the first file
df_first <- read.csv("./data5/Rare-Earths.csv") # Rare-Earths


# Read the lookup file
df_lookup <- read_excel("./data4/lookup.xlsx")
#   Country year  percentage
# 1      AE 1996 0.002591378
# 2      AE 1997 0.000633066
# 3      AE 1999 0.000608049
# 4      AE 2001 0.003855382
# 5      AE 2002 0.000841470
# 6      AE 2003 0.001065979


# For convenience, rename columns to simpler names
# Make sure these match the actual column names in your file
df_lookup <- df_lookup %>%
  rename(
    alpha2 = `Alpha-2`,
    alpha3 = `Alpha-3`,
    country_name = `Country`
  )

# A helper function that splits a code string (e.g. "AE-AT-EG")
# and looks up each Alpha-2 code in df_lookup
### ATTENTION
#### we need to add a column with "source = file_name"
convert_codes <- function(code_str) {
  codes <- str_split(code_str, "-", simplify = TRUE)
  codes <- codes[codes != ""]  # remove any empty parts
  
  iso3_list <- c()
  name_list <- c()
  
  for (c in codes) {
    # Find the matching row
    match_row <- df_lookup %>% filter(alpha2 == c)
    if (nrow(match_row) == 1) {
      iso3_list <- c(iso3_list, match_row$alpha3)
      name_list <- c(name_list, match_row$country_name)
    } else {
      iso3_list <- c(iso3_list, "???")
      name_list <- c(name_list, "Unknown")
    }
  }
  
  iso3_str <- paste(iso3_list, collapse = "-")
  country_names_str <- paste(name_list, collapse = ", ")
  
  return(list(iso3c = iso3_str, country_name = country_names_str))
}

# Vectorize the function so we can apply it easily
convert_codes_vec <- Vectorize(convert_codes, SIMPLIFY = FALSE)

# Apply it to the "Country" column in df_first
results <- convert_codes_vec(df_first$Country)

# results is a list of lists; extract iso3c and country_name into separate vectors
df_first$iso3c <- sapply(results, function(x) x$iso3c)
df_first$country_name <- sapply(results, function(x) x$country_name)

# Write out to Excel
write_xlsx(df_first, "./data6/Rare-Earths.xlsx")
tictoc::toc()
```


### Merge

```{r}
tictoc::tic()
library(tidyverse)

df.all <- read.csv("./data6/All.csv")
df.organic <- read.csv("./data6/Organic.csv")
df.organometallic <- read.csv("./data6/Organometallic.csv")
df.rare.earths <- read.csv("./data6/Rare-Earths.csv")


#put all data frames into list
df_list <- list(df.all, df.organic, df.organometallic, df.rare.earths)      

#merge all data frames together
df_merged <- df_list %>% reduce(full_join, by = c("iso2c", "year", "percentage", "iso3c", "country", "chemical")) |> 
  select(-X)

glimpse(df_merged)

write.csv(df_merged, "./data7/df_0.99.csv") 

tictoc::toc()

```



## Add information




```{r}

df <- read.csv("./data7/df_0.99.csv") |> 
  dplyr::select(
    -country
  )
  # dplyr::mutate(
  #   region = countrycode::countrycode(
  #     "ìso3c", 
  #     origin = "iso3c", 
  #     destination = "continent"
  #     )
  #     )

df.old <- read.csv("./data7/old.csv") 
  # rename(
  #   region = "region_23"
  # )


El.Df <- df.old |> 
  dplyr::left_join(df, by = c("year", "iso3c", "chemical")) |> 
  tidyr::drop_na(percentage)

write.csv(El.Df, "./data7/df.csv")

explore::explore(El.Df)

```

```{r}
sort(unique(df$year))
sort(unique(df.old$year))

unique(df$iso3c)
unique(df.old$iso3c)

unique(df$chemical)
unique(df.old$chemical)

unique(El.Df$region)
```